---
title: "Lecture 29: Delta method and variance stabilizing transformations"
author: "Ciaran Evans"
output: beamer_presentation
---

## Motivating example: Exponential confidence interval

$$X_1,...,X_n \overset{iid}{\sim} Exponential(\theta)$$

**Last time:** Used a pivotal quantity to find $1 - \alpha$ CI for $\theta$:

$$\left[ \frac{a}{\sum_i X_i}, \frac{b}{\sum_i X_i} \right]$$

**Alternative:** Wald Confidence interval

* MLE: $\widehat{\theta} =$

\bigskip

* Asymptotic distribution: $\sqrt{n}(\widehat{\theta} - \theta) \approx$

\bigskip

* Wald CI:

\vspace{2cm}

## Motivating example: non-constant variance

**Example:** Data on the number of hospitals and number of doctors (MDs) in US counties

```{r echo=F, message=F, warning=F, fig.width=3.5, fig.height=2.5}
library(tidyverse)
library(Stat2Data)
data("CountyHealth")

CountyHealth |>
  ggplot(aes(x = Hospitals, y = MDs)) +
  geom_point(size=1) +
  theme_bw()
```

**Question:** How do we adjust for non-constant variance in a linear model?

## Motivating example: non-constant variance

**Example:** Data on the number of hospitals and number of doctors (MDs) in US counties

```{r echo=F, message=F, warning=F, fig.width=3.5, fig.height=2.5}
CountyHealth |>
  ggplot(aes(x = Hospitals, y = sqrt(MDs))) +
  geom_point(size=1) +
  theme_bw()
```

## Goal: variance stabilizing transformation

Suppose $\widehat{\theta}$ is an estimator, and $Var(\widehat{\theta})$ depends on $\theta$. Examples:

* Exponential: $\sqrt{n}(\widehat{\theta} - \theta) \approx N(0, \theta^2)$

\bigskip

* Poisson: $\sqrt{n}(\widehat{\lambda} - \lambda) \approx N(0, \lambda)$

\bigskip

* Bernoulli: $\sqrt{n}(\widehat{p} - p) \approx N(0, p(1-p))$

\bigskip

**Goal:** Find a transformation $g$ such that $Var(g(\widehat{\theta}))$ does **not** depend on $\theta$

## Delta method

## Delta method

Suppose that $\widehat{\theta}$ is an estimate of $\theta \in \mathbb{R}$, such that 

$$\sqrt{n}(\widehat{\theta} - \theta) \overset{d}{\to} N(0, \sigma^2)$$

for some $\sigma^2$, and $g$ is a continuously differentiable function with $g'(\theta) \neq 0$. Then

$$\sqrt{n}(g(\widehat{\theta}) - g(\theta)) \overset{d}{\to} N(0, \sigma^2[g'(\theta)]^2)$$

\vspace{3cm}

## Example: Exponential

$$\sqrt{n}(\widehat{\theta} - \theta) \approx N(0, \theta^2)$$

\vspace{5cm}

## Example: Poisson

$$\sqrt{n}(\widehat{\lambda} - \lambda) \approx N(0, \lambda)$$

\vspace{5cm}

## Example: Bernoulli

$$\sqrt{n}(\widehat{p} - p) \approx N(0, p(1-p))$$

\vspace{5cm}


## Activity

Verify in simulations that a variance stabilizing transformation works (produces intervals with the desired coverage):

[https://sta711-s25.github.io/class_activities/ca_lecture_29.html](https://sta711-s25.github.io/class_activities/ca_lecture_29.html)



