---
title: "Lecture 9: Inference with logistic regression models"
author: "Ciaran Evans"
output: beamer_presentation
---

## Recall: the Titanic data

```{r, include=F}
titanic <- read.csv("https://sta711-s24.github.io/homework/Titanic.csv")
```

Data on 891 passengers on the *Titanic*. Variables include:

* `Survived`
* `Pclass`
* `Sex`
* `Age`

$$Survived_i \sim Bernoulli(p_i)$$

\vspace{-1cm}

$$\log \left( \frac{p_i}{1 - p_i} \right) = \beta_0 + \beta_1 Male_i + \beta_2 Age_i + \beta_3 Class2_i + \beta_4 Class3_i$$

## Fitting the model in R

```{r, include=F}
m1 <- glm(Survived ~ as.factor(Pclass) + Sex + Age, 
          family = binomial, data = titanic)
summary(m1)$coefficients
```

```
           Estimate Std. Error z value     Pr(>|z|)
(Intercept)   3.777      0.401   9.416    4.682e-21
Sexmale      -2.523      0.207 -12.164    4.811e-34
Age          -0.037      0.008  -4.831    1.359e-06
Pclass2      -1.310      0.278  -4.710    2.472e-06
Pclass3      -2.581      0.281  -9.169    4.761e-20
```

Suppose I want to know whether there is a relation between age and the probability of survival, after accounting for passenger class and sex. What hypotheses would I test?

\vspace{5cm}

## Wald tests for single coefficients

```
           Estimate Std. Error z value     Pr(>|z|)
(Intercept)   3.777      0.401   9.416    4.682e-21
Sexmale      -2.523      0.207 -12.164    4.811e-34
Age          -0.037      0.008  -4.831    1.359e-06
Pclass2      -1.310      0.278  -4.710    2.472e-06
Pclass3      -2.581      0.281  -9.169    4.761e-20
```

\vspace{6cm}

## Another question

```
           Estimate Std. Error z value     Pr(>|z|)
(Intercept)   3.777      0.401   9.416    4.682e-21
Sexmale      -2.523      0.207 -12.164    4.811e-34
Age          -0.037      0.008  -4.831    1.359e-06
Pclass2      -1.310      0.278  -4.710    2.472e-06
Pclass3      -2.581      0.281  -9.169    4.761e-20
```

Suppose I want to know whether there is a relation between *passenger class* and the probability of survival, after accounting for age and sex. What hypotheses would I test?

\vspace{5cm}

## Nested models

$$Survived_i \sim Bernoulli(p_i)$$

**Full model:**

\vspace{-0.5cm}

$$\log \left( \frac{p_i}{1 - p_i} \right) = \beta_0 + \beta_1 Male_i + \beta_2 Age_i + \beta_3 Class2_i + \beta_4 Class3_i$$

**Hypotheses:** 

\vspace{-0.5cm}

$$H_0: \beta_3 = \beta_4 = 0 \hspace{1cm} H_A: \text{ at least one of } \beta_3, \beta_4 \neq 0$$

**Reduced model:**

\vspace{-0.5cm}

$$\log \left( \frac{p_i}{1 - p_i} \right) = \beta_0 + \beta_1 Male_i + \beta_2 Age_i$$

## Logistic regression model performance

```
Coefficients:
                Estimate Std. Error z value Pr(>|z|)    
(Intercept)     3.777013   0.401123   9.416  < 2e-16 ***
Sexmale        -2.522781   0.207391 -12.164  < 2e-16 ***
Age            -0.036985   0.007656  -4.831 1.36e-06 ***
Pclass2        -1.309799   0.278066  -4.710 2.47e-06 ***
Pclass3        -2.580625   0.281442  -9.169  < 2e-16 ***
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

(Dispersion parameter for binomial family taken to be 1)

    Null deviance: 964.52  on 713  degrees of freedom
Residual deviance: 647.28  on 709  degrees of freedom
  (177 observations deleted due to missingness)
AIC: 657.28

Number of Fisher Scoring iterations: 5
```

## Logistic regression model performance

```
    Null deviance: 964.52  on 713  degrees of freedom
Residual deviance: 647.28  on 709  degrees of freedom
  (177 observations deleted due to missingness)
AIC: 657.28
```

* For logistic regression, deviance = $-2 \log \text{Likelihood}$
* Smaller deviances suggest a better fit to the data
* We compare nested models by comparing their deviances

## Nested logistic regression models

```{r, echo=T}
m1 <- glm(Survived ~ as.factor(Pclass) + Sex + Age, 
          family = binomial, data = titanic)

m1$deviance
```

```{r, echo=T}
m2 <- glm(Survived ~ Sex + Age, 
          family = binomial, data = titanic)
m2$deviance
```

$H_0$: the larger model is not a better fit

**Test statistic:** $G = 2( \log L_{\text{full}} - \log L_{\text{reduced}})$

\vspace{2cm}


## Nested logistic regression models

**Distribution:** Under $H_0$, $G \sim \chi^2_q$

* $q =$ difference in number of parameters

\vspace{1cm}

```{r, echo=T}
m1 <- glm(Survived ~ as.factor(Pclass) + Sex + Age, 
          family = binomial, data = titanic)


m2 <- glm(Survived ~ Sex + Age, 
          family = binomial, data = titanic)

pchisq(m2$deviance - m1$deviance, df=2, lower.tail=F)
```