---
title: "Lecture 2: Fitting and interpreting logistic regression models"
author: "Ciaran Evans"
output: beamer_presentation
---


```{r, include=F}
library(knitr)
library(tidyverse)

dengue <- read.csv("https://sta279-s22.github.io/labs/dengue.csv")
```

## Last time: Dengue data

**Data:** Data on 5720 Vietnamese children, admitted to the hospital with possible dengue fever. Variables include:

* *Sex*: patient's sex (female or male)
* *Age*: patient's age (in years)
* *WBC*: white blood cell count
* *PLT*: platelet count
* other diagnostic variables...
* *Dengue*: whether the patient has dengue (0 = no, 1 = yes)


## Logistic regression model

$$Y_i \sim Bernoulli(p_i)$$

$$\log \left(\dfrac{p_i}{1 - p_i}\right) = \beta_0 + \beta_1 WBC_i$$

Why is there no noise term $\varepsilon_i$ in the logistic regression model? Discuss for 1--2 minutes with your neighbor, then we will discuss as a class.

\vspace{4cm}

## Fitting the logistic regression model

$$Y_i \sim Bernoulli(p_i)$$

$$\log \left(\dfrac{p_i}{1 - p_i}\right) = \beta_0 + \beta_1 WBC_i$$

```{r, eval=F, echo=T}
m1 <- glm(Dengue ~ WBC, data = dengue, 
          family = binomial)
summary(m1)
```

\vspace{2cm}


## Fitting the logistic regression model

$$Y_i \sim Bernoulli(p_i)$$

$$\log \left(\dfrac{p_i}{1 - p_i}\right) = \beta_0 + \beta_1 WBC_i$$

```{r, output.lines=9:13}
m1 <- glm(Dengue ~ WBC, data = dengue, 
          family = binomial)
summary(m1)
```


## Making predictions

$$\log \left(\dfrac{\widehat{p}_i}{1 - \widehat{p}_i}\right) = 1.737 - 0.361 \ WBC_i$$

Work in groups of 2-3 on the following questions:

* What is the predicted odds of dengue for a patient with a WBC of 10?
* For a patient with a WBC of 10, is the predicted probability of dengue $> 0.5$, $< 0.5$, or $= 0.5$?
* What is the predicted *probability* of dengue for a patient with a WBC of 10?


## Shape of the regression curve

:::: {.columns}

::: {.column width="50%"}
$\log \left( \dfrac{p_i}{1 - p_i} \right) = \beta_0 + \beta_1 \ X_i \hspace{0.5cm}$

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=2.3, fig.height=2}

data.frame(x = seq(-5, 6, length.out=100),
           y1 = -1 + seq(-5, 6, length.out=100)) %>%
  ggplot(aes(x = x, y = y1)) +
  geom_line() +
  theme_bw() +
  labs(x = "x", y = "Log odds") 
```
:::

::: {.column width="50%"}
$p_i = \dfrac{e^{\beta_0 + \beta_1 \ X_i}}{1 + e^{\beta_0 + \beta_1 \ X_i}}$

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=2.3, fig.height=2}

expit <- function(p){
  return(exp(p)/(1 + exp(p)))
}

data.frame(x = seq(-5, 6, length.out=100),
           y1 = expit(-1 + seq(-5, 6, length.out=100))) %>%
  ggplot(aes(x = x, y = y1)) +
  geom_line() +
  theme_bw() +
  labs(x = "x", y = "P(y = 1)")
```
:::

::::



## Plotting the fitted model for dengue data

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=4, fig.height=3}
dengue %>%
  ggplot(aes(x = WBC, y = Dengue)) +
  geom_point() +
  geom_smooth(method = "glm", method.args = list(family = "binomial"), se=F) +
  theme_bw() +
  labs(y = "P(Dengue = 1 | WBC)")
```


## Shape of the regression curve

How does the shape of the fitted logistic regression depend on $\beta_0$ and $\beta_1$?

:::: {.columns}

::: {.column width="50%"}
$p_i = \dfrac{\exp\{\beta_0 +  X_i \}}{1 + \exp\{\beta_0 + X_i \}}$ for $\beta_0 = -3, -1, 1$

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=2.3, fig.height=2}

expit <- function(p){
  return(exp(p)/(1 + exp(p)))
}

data.frame(x = seq(-5, 6, length.out=100),
           y1 = expit(-1 + seq(-5, 6, length.out=100)),
           y2 = expit(-3 + seq(-5, 6, length.out=100)),
           y3 = expit(1 + seq(-5, 6, length.out=100))) %>%
  ggplot(aes(x = x, y = y1)) +
  geom_line() +
  geom_line(aes(y = y2), lty = 2, color="blue") +
  geom_line(aes(y = y3), lty = 3, color="red") +
  theme_bw() +
  labs(x = "x", y = "P(y = 1)") +
  annotate("text", x = -2, y = 0.4, label="1", size=3) +
  annotate("text", x = 0, y = 0.4, label="-1", size=3) +
  annotate("text", x = 3.5, y = 0.4, label="-3", size=3)
```
:::

::: {.column width="50%"}
$p_i = \dfrac{\exp\{-1 +  \beta_1 \ X_i \}}{1 + \exp\{-1 +  \beta_1 \ X_i \}}$ for $\beta_1 = 0.5, 1, 2$

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=2.3, fig.height=2}

data.frame(x = seq(-5, 6, length.out=100),
           y1 = expit(-1 + 0.5*seq(-5, 6, length.out=100)),
           y2 = expit(-1 + seq(-5, 6, length.out=100)),
           y3 = expit(-1 + 2*seq(-5, 6, length.out=100))) %>%
  ggplot(aes(x = x, y = y1)) +
  geom_line() +
  geom_line(aes(y = y2), lty = 2, color="blue") +
  geom_line(aes(y = y3), lty = 3, color="red") +
  theme_bw() +
  labs(x = "x", y = "P(y = 1)") +
  annotate("text", x = 3, y = 0.75, label="0.5", size=3) +
  annotate("text", x = 1.5, y = 0.75, label="1", size=3) +
  annotate("text", x = 0.5, y = 0.75, label="2", size=3) 
```
:::

::::



## Interpretation

$$\log \left(\dfrac{\widehat{p}_i}{1 - \widehat{p}_i}\right) = 1.737 - 0.361 \ WBC_i$$

Work in groups of 2-3 for on the following questions:

* Are patients with a higher WBC more or less likely to have dengue?
* What is the change in *log odds* associated with a unit increase in WBC?
* What is the change in *odds* asociated with a unit increase in WBC?