---
title: "Lecture 25: Likelihood ratio tests"
author: "Ciaran Evans"
output: beamer_presentation
---

```{r include=F}
library(knitr)
library(tidyverse)
library(MASS)
library(foreign)
library(statmod)
library(gridExtra)

earthquake <- read.csv("https://sta711-s23.github.io/homework/earthquake_small.csv")

```

## Asymptotics of the LRT

Suppose we observe iid data $X_1,...,X_n$ from a distribution with parameter $\theta \in \mathbb{R}$, and we wish to test $H_0: \theta = \theta_0$ vs. $H_A: \theta \neq \theta_0$. 

**Theorem:** Under $H_0$ (and assuming required regularity conditions),

$$2 \log \left( \frac{L(\widehat{\theta}_{MLE} | {\bf X})}{L(\theta_0 | {\bf X})} \right) \overset{d}{\to} \chi^2_1$$

\vspace{4cm}


## Generalization to higher dimensions

Suppose we observe iid data $X_1,...,X_n$ with parameter $\theta \in \mathbb{R}^d$. Partition $\theta = (\theta_{(1)}, \theta_{(2)})^T$, with $\theta_{(2)} \in \mathbb{R}^q$. We wish to test
$$H_0: \theta_{(2)} = {\bf 0} \hspace{1cm} H_A: \theta_{(2)} \neq {\bf 0}$$

**Theorem:** Under $H_0$ (and assuming required regularity conditions), 

$$2 \log \left( \frac{\sup_\theta L(\theta | {\bf X})}{\sup \limits_{\theta: \theta_{(2)} = 0} L(\theta | {\bf X})} \right) \overset{d}{\to} \chi^2_q$$

\vspace{4cm}

## Earthquake data

Data from the 2015 Gorkha earthquake on 211774 buildings, with variables including:

* `Damage`: whether the building sustained any damage (1) or not (0)
* Age: the age of the building (in years)
* `Surface`: a categorical variable recording the surface condition of the land around the building. There are three different levels: n, o, and t


## Likelihood ratio tests

```{r, echo=T}
m1 <- glm(Damage ~ Age*Surface, data = earthquake,
          family = binomial)
summary(m1)$coefficients
```

We want to test whether the relationship between Age and Damage is the same for all three surface conditions. What hypotheses do we test?


## Likelihood ratio tests

**Full model:**

```{r, eval=F, echo=T}
m1 <- glm(Damage ~ Age*Surface, data = earthquake,
          family = binomial)
```

**Reduced model:**

```{r, eval=F, echo=T}
m2 <- glm(Damage ~ Age + Surface, data = earthquake,
          family = binomial)
```


## Comparing deviances

```{r, echo=T}
m1 <- glm(Damage ~ Age*Surface, data = earthquake,
          family = binomial)
m1$deviance

m2 <- glm(Damage ~ Age + Surface, data = earthquake,
          family = binomial)
m2$deviance
```

## Comparing deviances

```{r, echo=T}
m1 <- glm(Damage ~ Age*Surface, data = earthquake,
          family = binomial)

m2 <- glm(Damage ~ Age + Surface, data = earthquake,
          family = binomial)

pchisq(m2$deviance - m1$deviance,
       m2$df.residual - m1$df.residual,
       lower.tail = F)
```

