---
title: "Lecture 16: Wald tests"
author: "Ciaran Evans"
output: beamer_presentation
---

## Wald test for one parameter

## Wald test for one parameter: examples

* Population mean: $H_0: \mu = \mu_0$

\vspace{1.5cm}

* Population proportion: $H_0: p= p_0$

\vspace{1.5cm}

* Regression coefficient: $H_0: \beta_j = 0$

\vspace{2cm}

## Testing multiple parameters

Logistic regression model for the dengue data:

$$Y_i \sim Bernoulli(p_i)$$

$$\log \left( \dfrac{p_i}{1 - p_i} \right) = \beta_0 + \beta_1 WBC_i + \beta_2 PLT_i$$

Researchers want to know if there is any relationship between white blood cell count or platelet count, and the probability a patient has dengue. 

**Question:** What hypotheses should they test?

\vspace{2cm}

## Testing multiple parameters

```{r include=F}
library(knitr)


dengue <- read.csv("https://sta712-f22.github.io/homework/dengue.csv")
```

```{r, echo=F}
m1 <- glm(Dengue ~ WBC + PLT, data = dengue, 
          family = binomial)
summary(m1)$coefficients |> round(digits=3)
```

$H_0: \beta_1 = \beta_2 =0$

Can the researchers test their hypotheses using this output?

\vspace{4cm}

## Wald tests for multiple parameters

For the dengue example:

$$\widehat{\beta} = \begin{pmatrix} \widehat{\beta}_0 \\ \widehat{\beta}_1 \\ \widehat{\beta}_2 \end{pmatrix} \hspace{3cm}$$

```{r}
m1 <- glm(Dengue ~ WBC + PLT, data = dengue, family = binomial)
coef(m1)
vcov(m1)
```

## Multivariate normal distribution

**Definition:** Let $X = (X_1,...,X_k)^T$. We say that $X \sim N(\mu, \Sigma)$ if for any ${\bf a} \in \mathbb{R}^k$, ${\bf a}^T X$ follows a (univariate) normal distribution.

* $\mu =$

\vspace{2cm}

* $\Sigma =$

\vspace{2cm}

## Multivariate normal distribution

```{r, echo=F, message=F, warning=F, fig.align='center', fig.width=3.5, fig.height=3.5}
library(MASS)
library(patchwork)
library(tidyverse)
y <- mvrnorm(1000, mu=c(0, 0), Sigma = diag(2))

p1 <- data.frame(x1 = y[,1], x2 = y[,2]) |>
  ggplot(aes(x = x1, y = x2)) +
  geom_point(alpha = 0.2, size=0.3) +
  stat_ellipse(type="norm", level = 0.5) +
  stat_ellipse(type="norm", level = 0.7) +
  stat_ellipse(type="norm", level = 0.9) +
  theme_classic()

y <- mvrnorm(1000, mu=c(0, 0), Sigma = diag(c(1, 5)))

p2 <- data.frame(x1 = y[,1], x2 = y[,2]) |>
  ggplot(aes(x = x1, y = x2)) +
  geom_point(alpha = 0.2, size=0.3) +
  stat_ellipse(type="norm", level = 0.5) +
  stat_ellipse(type="norm", level = 0.7) +
  stat_ellipse(type="norm", level = 0.9) +
  theme_classic() +
  scale_x_continuous(limits = c(-6, 6)) +
  scale_y_continuous(limits = c(-6, 6))

y <- mvrnorm(1000, mu=c(0, 0), Sigma = matrix(c(1, 0.75, 0.75, 1), nrow=2))

p3 <- data.frame(x1 = y[,1], x2 = y[,2]) |>
  ggplot(aes(x = x1, y = x2)) +
  geom_point(alpha = 0.2, size=0.3) +
  stat_ellipse(type="norm", level = 0.5) +
  stat_ellipse(type="norm", level = 0.7) +
  stat_ellipse(type="norm", level = 0.9) +
  theme_classic() +
  scale_x_continuous(limits = c(-3, 3)) +
  scale_y_continuous(limits = c(-3, 3))

(p1 + p2)/(p3 + plot_spacer())

#alpha_levels <- c(0.5, 0.9)
#contour_data <- data.table::rbindlist(lapply(alpha_levels,ellipse,x=diag(2),
#      scale=c(1,1),  ## needed for positional matching
#      centre=c(0, 0)))
```

## Wald tests for multiple parameters

For the dengue example: $\widehat{\beta} = \begin{pmatrix} \widehat{\beta}_0 \\ \widehat{\beta}_1 \\ \widehat{\beta}_2 \end{pmatrix}$

We want to test: $\begin{pmatrix} \beta_1 \\ \beta_2 \end{pmatrix} = \begin{pmatrix} 0 \\ 0 \end{pmatrix}$

\vspace{5cm}

## Wald tests for multiple parameters

$$H_0: {\bf C} \beta = \gamma_0$$
For the dengue example:

```{r}
C <- matrix(c(0, 1, 0,
              0, 0, 1), nrow=2, byrow=T)

C
C %*% coef(m1)
```

## Wald tests for multiple parameters

For the dengue example:

```{r}
C
vcov(m1)
C %*% vcov(m1) %*% t(C)
```



## Wald tests for multiple parameters

* $H_0: {\bf C} \beta = \gamma_0$

* Look at ${\bf C} \widehat{\beta}$

**Fact:** Suppose $X \sim N(\mu, \Sigma)$ (multivariate normal), and ${\bf A}$ is a matrix (not random). Then:

$${\bf A} X \sim \hspace{3cm}$$

\vspace{3cm}

## Test statistic and p-value

* $H_0: {\bf C} \beta = \gamma_0$

* ${\bf C} \widehat{\beta} \approx N({\bf C} \beta, \ {\bf C} \mathcal{I}^{-1}(\beta){\bf C}^T)$

* Want to turn ${\bf C} \widehat{\beta}$ into a scalar test statistic

\vspace{5cm}

## Example

$H_0: {\bf C} \beta = \gamma_0$

\vspace{-0.5cm}

$$({\bf C} \widehat{\beta} - \gamma_0)^T ({\bf C} \mathcal{I}^{-1}(\beta) {\bf C}^T)^{-1} ({\bf C} \widehat{\beta} - \gamma_0) \approx \chi^2_q \hspace{1cm} \text{under } H_0$$

```{r, eval=F}
C %*% coef(m1)
```

```{r, echo=F}
c(C %*% coef(m1))
```



```{r}
test_stat <- t(C %*% coef(m1)) %*% solve(C %*% vcov(m1) %*% t(C)) %*%
  (C %*% coef(m1))
test_stat
pchisq(test_stat, 2, lower.tail=F)
```




