---
title: "Lecture 27: Interval estimation"
author: "Ciaran Evans"
output: beamer_presentation
---

## Motivation

Suppose we have data $(X_1, Y_1),...,(X_n, Y_n)$ with
$$Y_i \sim Bernoulli(p_i)$$
$$\log \left( \dfrac{p_i}{1 - p_i} \right) = \beta^T X_i$$

So far, we have discussed:

* Finding point estimates $\widehat{\beta}$
* Testing hypotheses about the true (but unknown) parameters $\beta$

**Question:** What are the limitations of point estimates and hypothesis tests for inference about $\beta$?

## Confidence interval

```{r, include=F}
dengue <- read.csv("https://sta712-f22.github.io/homework/dengue.csv")
```

```{r, echo=F}
m1 <- glm(Dengue ~ WBC + PLT, data = dengue, 
          family = binomial)
summary(m1)$coefficients
```

**Question:** How would I calculate a 95% confidence interval for $\beta_1$ (the change in the log odds of dengue for a one-unit increase in WBC, holding PLT fixed)?

\vspace{4cm}

## Confidence interval

```{r, echo=F}
m1 <- glm(Dengue ~ WBC + PLT, data = dengue, 
          family = binomial)
summary(m1)$coefficients
```

95% confidence interval for $\beta_1$: (-0.315, -0.262)

**Question:** How do I interpret this confidence interval?

\vspace{4cm}


## Deriving the coverage probability

$$(1 - \alpha) \text{ Wald interval}: \widehat{\theta} \pm z_{\alpha/2} SE(\widehat{\theta})$$

\vspace{6cm}

## Formal definition

## Inverting a test

**Theorem:** Let $\theta \in \Theta$ be a parameter of interest. For each value of $\theta_0 \in \Theta$, consider testing $H_0: \theta = \theta_0$ vs. $H_A: \theta \neq \theta_0$, and let $\mathcal{R}(\theta_0)$ be the rejection region for a level $\alpha$ test. Let $C(X_1,...,X_n) = \{\theta_0 \in \Theta: (X_1,...,X_n) \not \in \mathcal{R}(\theta_0) \}$. Then $C(X_1,...,X_n)$ is a $1 - \alpha$ confidence set for $\theta$.

\vspace{4cm}

## Example

Suppose $X_1,...,X_n \overset{iid}{\sim} Uniform[0, \theta]$. Let's invert the LRT to find a confidence interval for $\theta$.

\vspace{6cm}





