---
title: "Class Activity Solutions"
format: 
  html:
    embed-resources: true
editor: source
author: "Ciaran Evans"
---

1.

```{r}
n <- 2
nsim <- 1000
means <- rep(NA, nsim)

for(i in 1:nsim){
  x <- rpois(n, 1)
  means[i] <- mean(x)
}

cdf_pts <- seq(-4, 4, 0.01)
plot(cdf_pts, ecdf(sqrt(n) * (means - 1) / sqrt(1))(cdf_pts), type="l", 
     xlab = "x", ylab = "CDF")
points(cdf_pts, pnorm(cdf_pts, mean=0, sd=1), type="l", col="red")
```

A normal distribution isn't a perfect approximation here; the distribution of $\overline{X}$ is discrete, and only tends to take a few different values. Nevertheless, the normal distribution does capture the general shape pretty well.

2.

```{r}
n <- 5
nsim <- 1000
means <- rep(NA, nsim)

for(i in 1:nsim){
  x <- rpois(n, 1)
  means[i] <- mean(x)
}

cdf_pts <- seq(-4, 4, 0.01)
plot(cdf_pts, ecdf(sqrt(n) * (means - 1) / sqrt(1))(cdf_pts), type="l", 
     xlab = "x", ylab = "CDF", main = "n = 5")
points(cdf_pts, pnorm(cdf_pts, mean=0, sd=1), type="l", col="red")
```

```{r}
n <- 10
nsim <- 1000
means <- rep(NA, nsim)

for(i in 1:nsim){
  x <- rpois(n, 1)
  means[i] <- mean(x)
}

cdf_pts <- seq(-4, 4, 0.01)
plot(cdf_pts, ecdf(sqrt(n) * (means - 1) / sqrt(1))(cdf_pts), type="l", 
     xlab = "x", ylab = "CDF", main = "n = 10")
points(cdf_pts, pnorm(cdf_pts, mean=0, sd=1), type="l", col="red")
```


```{r}
n <- 30
nsim <- 1000
means <- rep(NA, nsim)

for(i in 1:nsim){
  x <- rpois(n, 1)
  means[i] <- mean(x)
}

cdf_pts <- seq(-4, 4, 0.01)
plot(cdf_pts, ecdf(sqrt(n) * (means - 1) / sqrt(1))(cdf_pts), type="l", 
     xlab = "x", ylab = "CDF", main = "n = 30")
points(cdf_pts, pnorm(cdf_pts, mean=0, sd=1), type="l", col="red")
```

As $n$ increases, the normal distribution becomes a better approximation for the distribution of $\overline{X}$.

3. Here is an $Exponential(1)$ distribution; we can see the normal approximation still performs well:

```{r}
n <- 30
nsim <- 1000
means <- rep(NA, nsim)

for(i in 1:nsim){
  x <- rexp(n, 1)
  means[i] <- mean(x)
}

cdf_pts <- seq(-4, 4, 0.01)
plot(cdf_pts, ecdf(sqrt(n) * (means - 1) / sqrt(1))(cdf_pts), type="l", 
     xlab = "x", ylab = "CDF", main = "n = 30")
points(cdf_pts, pnorm(cdf_pts, mean=0, sd=1), type="l", col="red")
```



