---
title: "Class Activity"
output: 
  rmdformats::robobook:
    css: "homework.css"
    highlight: pygments
link-citations: yes
---

# Parameter estimation with a mis-specified model

Suppose that we observe $(X_i, Y_i)$ with $X_i \sim N(0, 0.25)$, $Y_i \sim Poisson(\mu_i)$, and 

$$\log(\mu_i) = -1 + 0.2 X_i^3$$
**However**, we incorrectly fit a Poisson model with $\log(\mu_i) = \beta_0 + \beta_1 X_i$ (i.e., our shape assumption is incorrect, even though we got the distribution right).

The code below simulates data from this true distribution, and fits the incorrectly-specified model:

```{r}
n <- 500
x <- rnorm(n, sd=0.5)

# true relationship (cubic)
lambda <- exp(-1 + 0.2*x^3)
y <- rpois(n, lambda)

# fitted model (incorrect shape assumption)
m1 <- glm(y ~ x, family = poisson)
```

As $n \to \infty$, we know that $\beta \overset{p}{\to} \beta^*$, and in the previous activity we estimated $\beta_1^*$ (which was approximately 0.155).

To verify that $\widehat{\beta} \approx N(\beta^*, {\bf S}_n)$, where ${\bf S}_n$ is the sandwich variance, we will show that confidence intervals constructed with the sandwich variance have the desired coverage for $\beta^*$.

The code below simulates data as described above, and calculates a 95% CI for $\beta_1^*$ using the sandwich variance estimate.

```{r, eval=F}
n <- 500

x <- rnorm(n, sd=0.5)
lambda <- exp(-1 + 0.2*x^3)
y <- rpois(n, lambda)

m1 <- glm(y ~ x, family = poisson)
cov_mat <- sandwich(m1)

upper <- m1$coefficients[2] + 1.96*sqrt(cov_mat[2,2])
lower <- m1$coefficients[2] - 1.96*sqrt(cov_mat[2,2])
```

:::{.question}
#### Question 1

Repeat the code many times, and verify that the confidence intervals do indeed contain $\beta_1^*$ approximately 95% of the time.
:::


## Using the sandwich variance with a *correctly* specified model

The simulations above (and in the prior class activity) show that the sandwich variance works with a mis-specified model. Importantly, we can also use the sandwich variance with a *correctly* specified model!

To show that the sandwich variance works with a correctly specified model, we will generate data from a Poisson model, fit a Poisson regression for which all assumptions are met, and calculate a confidence interval for $\beta_1$:

```{r, eval=F}
n <- 500

x <- rnorm(n)
lambda <- exp(1 + x)
y <- rpois(n, lambda)

m1 <- glm(y ~ x, family = poisson)
cov_mat <- sandwich(m1)

upper <- m1$coefficients[2] + 1.96*sqrt(cov_mat[2,2])
lower <- m1$coefficients[2] - 1.96*sqrt(cov_mat[2,2])
```

:::{.question}
#### Question 2

Repeat the code many times, and verify that the confidence intervals do indeed have approximately 95% coverage.

:::


